On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-06 main.tex
proof of correctness

Notice that we call \(ϵ̂\) the accepting function and \(δ̂\) the transition function; intuitively, a state \(s\) accepts an atom \(α\) if and only if there exists a \(b ∈ ϵ̂(s)\), s.t. \(α ≤ b\); similarly, the state \(s\) transitions to \(s'\) via atom \(α\) while executing \(p\) if and only if there exists \((b, s', p) ∈ δ̂(s)\) and \(α ≤ b\).
With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ (S, ϵ̂, δ̂)\) can be lowered into a GKAT coalgebra \((S, δ)\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\[5px]
    (s', p) & 
        \begin{aligned}
            & ∃ b ∈ \BExp_B, α ≤ b \\
            & \text{ and } δ(s, b) = (s', p)  
        \end{aligned}\\[5px]
    \reject & \text{otherwise}
\end{cases}
\end{align}
This is well-defined, i.e. no more than one clause can be satisfied precisely because the boolean expressions appear in \(ϵ̂\) and \(δ̂\) are disjoint.
We usually use \(S\) to denote the lowering of \(Ŝ\).

@@

\begin{theorem}[Functoriality]
    The lowering operation is a functor, given a symbolic GKAT coalgebra homomorphism \(h: Ŝ → Û\), then \(h\) is also a homomorphism \(h: S → U\).
\end{theorem}

@@

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the least symbolic GKAT coalgebra (ordered by point-wise subset ordering on \(ϵ̂\) and \(δ̂\)) that satisfy the rules in~\Cref{fig:derivatives-rules}.
Notice that the rules listed on~\Cref{fig:derivatives-rules} is very close to that of~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
This is no coincidence, as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis on our correctness argument.
\begin{theorem}[Correctness]
    The lowering of \(D̂\) is exactly the derivative defined by Schmid et. al.~\cite{schmid_GuardedKleeneAlgebra_2021}.
    Therefore, the semantics of the expression is equal to the semantics \(⟦e⟧ = ⟦e⟧_{D̂}\).
    TODO: unfold the statement.
\end{theorem}

@@

>>>>

Notice that we call \(ϵ̂\) the accepting function and \(δ̂\) the transition function; intuitively, a state \(s\) accepts an atom \(α\) if and only if there exists a \(b ∈ ϵ̂(s)\), s.t. \(α ≤ b\); similarly, the state \(s\) transitions to \(s'\) via atom \(α\) while executing \(p\) if and only if there exists \((b, s', p) ∈ δ̂(s)\) and \(α ≤ b\).
With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ (S, ϵ̂, δ̂)\) can be lowered into a GKAT coalgebra \((S, δ)\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\[5px]
    (s', p) & 
        \begin{aligned}
            & ∃ b ∈ \BExp_B, α ≤ b \\
            & \text{ and } δ(s, b) = (s', p)  
        \end{aligned}\\[5px]
    \reject & \text{otherwise}
\end{cases}
\end{align}
This is well-defined, i.e. no more than one clause can be satisfied precisely because the boolean expressions appear in \(ϵ̂\) and \(δ̂\) are disjoint.
We usually use \(S\) to denote the lowering of \(Ŝ\); and the semantics of a state \(s ∈ Ŝ\) is defined as its semantics in the lowering \(⟦s⟧_{Ŝ} ≜ ⟦s⟧_{S}.\)

@@

\begin{theorem}[Functoriality]\label{thm:lowering-functor}
    The lowering operation is a functor, given a symbolic GKAT coalgebra homomorphism \(h: Ŝ → Û\), then \(h\) is also a homomorphism \(h: S → U\).
\end{theorem}

@@

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the least symbolic GKAT coalgebra (ordered by point-wise subset ordering on \(ϵ̂\) and \(δ̂\)) that satisfy the rules in~\Cref{fig:derivatives-rules}.
Notice that the rules listed on~\Cref{fig:derivatives-rules} is very close to that of~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
This is no coincidence, as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis on our correctness argument.
\begin{theorem}[Correctness]\label{thm:derivative-correctness}
    The lowering of \(D̂\), denoted \(D\), is exactly the derivative defined by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
    Therefore, the semantics of the expression is equal to the semantics generated by the derivative coalgebra, \(⟦e⟧ = ⟦e⟧_{D} = ⟦e⟧_{D̂}\).
\end{theorem}
\begin{proof}
    TODO: unfold the statement.
\end{proof}

@@

\begin{proof}
    By functoriality of lowering~\cref{thm:lowering-functor}, then \(h: S_{e} → D\) is a homomorphism on their lowerings. 
    By standard theorem of universal coalgebra, homomorphism preserves the result to the final coalgebra~\cite{rutten_UniversalCoalgebraTheory_2000}, in particular, \(⟦s^*⟧^{ω}_{S_{e}} = ⟦h(s^*)⟧^{ω}_{D} = ⟦e⟧^{ω}_{D}\), where \(⟦-⟧^{ω}\) is the infinite trace semantics i.e. the unique map into the final GKAT coalgebra.

    Finally, because infinite trace equivalence implies finite trace equivalence (\Cref{thm:inf-trace-equiv-implies-fin-trace-equiv}), \(⟦s^*⟧_{S_{e}} = ⟦e⟧_{D}\); then the correctness of derivative (\Cref{thm:derivative-correctness}) concludes our proof: \(⟦s^*⟧_{S_{e}} = ⟦e⟧_{D} = ⟦e⟧\).
\end{proof}
